---
title: "Data Frames, Basic Indexing, Reading/Writing Data"
classdate: "9/2/2020 -- In Class"
teaching: 90
exercises: 30
questions:
- "Class questions."
objectives:
- "Class objectives."
keypoints:
- "Class keypoints."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-class-data-frames-")
```

***
## In Class

***
### Lists

Lists in R act as containers. Unlike atomic vectors, the contents of a list are
not restricted to a single mode and can encompass any mixture of data
types. Lists are sometimes called "generic vectors"", because the elements of a
list can by of any type of R object, even lists containing further lists. This
property makes them fundamentally different from atomic vectors.

A list is a special type of vector. Each element can be a different type, and is not restricted to a single value.

Create lists using `list()` or coerce other objects using `as.list()`. An empty
list of the required length can be created using `vector()`

```{r}
x <- list(1, "a", TRUE, 1+4i)
x # A list does not print to the console like a vector. Instead, each element of the list starts on a new line.

x <- vector("list", length = 5) # empty list
length(x)
x 
```

&nbsp;

Vectors can be coerced to lists as follows:

```{r}
x <- 1:10
x <- as.list(x)
length(x)
```

&nbsp;
#### Indexing lists

Indexing works a bit differently for lists. The content of elements of a list can be retrieved by using double square brackets `[[n]]`, as opposed to the single square brackets `[n]` used for vectors and matrices.

```{r}
x[[1]]
```


> ## Examining Lists
>
> 1. What is the class of `x[1]`?
> 2. What is the class of `x[[1]]`?
>
> > ## Solution
> > 1. 
> >     ```{r examine-lists-1}
> >     class(x[1])
> >     ```
> > 2. 
> >     ```{r examine-lists-2}
> >     class(x[[1]])
> >     ```
> {: .solution}
{: .challenge}

&nbsp;

Elements of a list can be named (i.e. lists can have the `names` attribute)

```{r}
xlist <- list(a = "Karthik Ram", b = 1:10, data = head(iris))
xlist
names(xlist)
attributes(xlist)
```

&nbsp;

You can use the `$` operator to directly refer to named list elements by their name:

```{r}
xlist[[1]]
xlist$a
```

> ## Examining Named Lists
>
> 1. What is the length of the `xlist` object?
> 2. What is its structure?
>
> > ## Solution
> > 1. 
> >     ```{r examine-named-lists-1}
> >     length(xlist)
> >     ```
> > 2. 
> >     ```{r examine-named-lists-2}
> >     str(xlist)
> >     ```
> {: .solution}
{: .challenge} 

&nbsp;
#### Lists of lists!

A list can even have vectors (or other lists!) as one element of a list:

```{r}
x <- 1:10
y <- c(T, F, T, T) # "T" can be used in place of "TRUE"; "F" can be used in place of "FALSE"
z <- list(1, "a", TRUE, 1+4i)

my.list <- list(x, y, z)
my.list

my.list[[1]]
class(my.list[[1]])
class(my.list[[2]])

my.list <- list(x = x, y = y, z = x) # use the `=` to name your list elements
my.list$x
```

&nbsp;
#### Lists and functions

Lists can be extremely useful inside functions. Because the functions in R are 
able to return only a single object, you can "staple" together lots of different
kinds of results into a single object that a function can return.

Elements are indexed by double brackets. Double brackets return the object inside the indexed list element, while single brackets will just return a(nother) list containing the indicated elements. 

```{r}
my.list[1]
my.list[[1]]
```


&nbsp;

Again, if the elements of a list are named, they can be referenced by
the `$` notation (i.e. `xlist$data`). This can be useful if a function runs a complex analysis. It can export a lot of different types of information as different elements in a single list (raw data, processes data, analysis summary, analysis statistics, etc.)


***
### Data Frames

The data frame is the most important data type in R. It is the *de facto*
data structure for most tabular data and what we use for statistics. At its heart, the data frame is a *special type of list* where every element of the list has same length (i.e. data frame is a "rectangular" list).

Data frames can have additional attributes such as `rownames()`, which can be
useful for annotating data, like `subject_id` or `sample_id`. But most of the
time they are not used.

Data frame characteristics:
* Each column must contain a single data type
* Different columns can have different data types
* All columns must have the same number of elements
* Like lists, column names can be directly called using the `$` operator
* Like matrices, data frames can be indexed using the `df[row,column]` notation.

Some additional information on data frames:

* Created by the `read.csv()` and `read.table()` when importing the data into R.
* Assuming all columns in a data frame are of same type, data frame can be converted to a matrix with data.matrix() (preferred) or as.matrix(). Otherwise type coercion will be enforced and the results may not always be what you expect.
* Can be created *de novo* with the `data.frame()` or `as.data.frame()` functions.
* Like matrices, the number of rows and columns can be queried with `nrow(dat)` and `ncol(dat)`, respectively.
* Rownames are often automatically generated and look like 1, 2, ..., n. Consistency in numbering of rownames may not be honored when rows are reshuffled or subset.
* A common way to use data frames is with columns as "variables" (e.g. body weight, blood pressure) and rows as "patients" or "observations" (e.g. "subject 1", "subject 2", ...).

The following table summarizes the one-dimensional and two-dimensional data structures in 
R in relation to diversity of data types they can contain.

| Dimensions | Homogenous | Heterogeneous |
| ------- | ---- | ---- |
| 1-D | atomic vector | list |
| 2-D | matrix | data frame |

&nbsp;
#### Creating Data Frames by Hand

To create data frames by hand:

```{r}
dat <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)
dat
```

&nbsp;

See that it is actually a special list:

```{r}
is.list(dat)
is.data.frame(dat) # "data.frame" is a sub-class of "list"
class(dat)
```

&nbsp;
#### Sample data in R

R includes a default dataset that can be used for examples caused `iris`. The `iris` data frame is already defined and loaded in memory. We will use this data frame to learn how to interact with data frames. 

Let's get some information on the `iris` data set:
```{r eval=FALSE}
?iris
```

&nbsp;
#### Useful function for examining and extracting information from data frames

There are many ways to interact with data frames and get information about their contents.

> ## Useful Data Frame Functions
>
> * `head()` - shows first 6 rows
> * `tail()` - shows last 6 rows
> * `dim()` - returns the dimensions of data frame (i.e. number of rows and number of columns)
> * `nrow()` - number of rows
> * `ncol()` - number of columns
> * `str()` - structure of data frame - name, type and preview of data in each column
> * `names()` or `colnames()` - both show the `names` attribute for a data frame
> * `sapply(dataframe, class)` - shows the class of each column in the data frame
{: .callout}

&nbsp;

I find `head()`, which displays the first 6 rows and all columns of the data frame, particularly useful to get a feel for the contents and organization of a new data frame. Let's take a quick look at `iris` with `head()` and `dim()`:

```{r}
head(iris)
head(iris, 2) # you can look at more or fewer rows if 6 is not what you want
dim(iris)
```

> ## Column Types in Data Frames
>
> Knowing that data frames are lists, can columns be of different type?
>
> What type of structure do you expect to see when you explore the structure of the `iris` data 
> frame? Hint: Use `str()`.
>
> > ## Solution
> > The Sepal.Length, Sepal.Width, Petal.Length and Petal.Width columns are all
> > numeric types, while Species is a Factor.
> > Lists can have elements of different types.
> > Since a Data Frame is just a special type of list, it can have columns of
> > differing type (although, remember that type must be consistent within each column!).
> > ```{r column-types}
> > str(iris)
> > ```
> {: .solution}
{: .challenge}

&nbsp;
#### Indexing data frames

R is a powerful language for data manipulation. There are three main ways for addressing data inside R objects.

* By index (subsetting)
* By name (columns only)
* By logical vector

&nbsp;

**By index**

Because data frames are rectangular, elements of data frame can be referenced by specifying 
the row and the column index in single square brackets (similar to matrix).

```{r}
iris[1, 3]
```

&nbsp;

Like matrices, we can also ask for multiple columns and rows using the `:` operator:
```{r}
iris[1:20, 2:3]
```

&nbsp;

Or for non-contiguous subsets using lists of indices:
```{r}
iris[c(1:10,20:25), c(1,3)] # note the use of sequences within the `c()` list
```

&nbsp;

What if we only ask for one dimension in data frame?

> ## Asking for a single index from a data frame
> 
> What do you expect if we query `iris` with just one index?
> 
> ```{r, eval=FALSE}
> iris[3]
> ```
> > ## Solution
> > ```{r}
> > iris[3]
> > ```
> > Because the data frame is a list at heart, just asking for a single index returns 
> > the indicated column.
> >
> {: .solution}
{: .challenge}

&nbsp;

What if we include the `,`, but don't give an actual index?

> ## Selecting More Values
>
> What will be returned by `dat[, 2]`?
>
> > ## Solution
> > ```{r select-values-2}
> > dat[, 2]
> > ```
> >
> > If we leave out a dimension R will interpret this as a request for all values in 
> > that dimension.
> {: .solution}
{: .challenge}

&nbsp;

**By name (columns only)**

As data frames are also lists, it is possible to refer to columns (which are elements of
such list) using the list notation, i.e. either double square brackets or a `$`.

```{r}
# These all return the same vector:
iris[[1]]
iris[["Sepal.Length"]]
iris$Sepal.Length
```

&nbsp;

We can ask for a set of columns using `c()`:

```{r}
iris[1:10, c("Sepal.Length","Petal.Length")]
```

&nbsp;

Named vectors can also be used in vectorized mathematical operations:

```{r}
iris$Sepal.Length + iris$Petal.Length
```

> ## Default Names
>
> If column names are not specified, `V1, V2, ..., Vn` are automatically used as the defaults. For example, if we convert a matrix to a data frame:
> 
> ```{r}
> m = matrix(1:10, nrow = 2)
> m.data = as.data.frame(m)
> m.data
> ``` 
{: .callout}

&nbsp;

**By logical vector**

Recall that a logical vector contains only the special values `TRUE` and `FALSE`.

```{r logical_vectors}
c(TRUE, TRUE, FALSE, FALSE, TRUE)
```

&nbsp;

Logical vectors can be created using `relational operators`:
* `<` = less than
* `>` = greater than
* `<=` = less than or equal to
* `>=` = greater than or equal to
* `==` = exactly equal to
* `!=` = not equal to
* `%in%` = is present in (to as if the value on the left is present in the vector/matrix on the right)

A few single variable examples:
```{r logical_variable_example}
1 == 1
1 == 2
1 != 1
4 > 7
18 %in% 1:10
18 %in% 15:25
```

&nbsp;

We can use these operators to query entire vectors and generate logical vectors:

```{r logical_vector_example}
# creating logical vectors from numeric data
x <- c(1, 2, 3, 11, 12, 13)
x < 10
x %in% 1:10
```

&nbsp;

We can use logical vectors to select data from a data frame.

```{r logical_vectors_indexing}
index <- iris$Species == 'setosa'
index
iris[index,]
```

&nbsp;

Often this operation is written as one line of code:

```{r logical_vectors_indexing2}
iris[iris$Species == 'setosa', ]
```

> ## Using logical indices
>
> Create a new data frame that is the subset of `iris` with sepal length greater than or
> equal to 5.0.
> 
> > ## Solution
> > ```{r}
> > iris.new = iris[iris$Sepal.Length > 5,]
> > ```
> {: .solution}
{: .challenge}

&nbsp;

In addition to the numeric comparisons, there are a set of operators that compare logical variables and output a new logical variable:
* `!` = NOT (changes `TRUE` to `FALSE` and vice versa)
* `&` = element-wise AND (both are true; outputs vector for vector input comparing elements)
* `&&` = logical AND (both are true; only considers first index of a vector)
* `|` = element-wise OR (one or both are true; outputs vector for vector input comparing elements)
* `&&` = logical OR (both are true; only considers first index of a vector)

```{r}
truth = c(TRUE, FALSE, TRUE, TRUE)
lie = !truth
truth
lie

T & T
T & F
T | F
F | F
```

***
### Exercises

> ## Using logical indices
>
> Create a new data frame that is the subset of `iris` with sepal length greater than or > equal to 5.0 for the setosa species.
> 
> > ## Solution
> > ```{r}
> > iris.new = iris[iris$Sepal.Length > 5 & iris$Species == "setosa",]
> > ```
> {: .solution}
{: .challenge}

> ## Subsetting using a vector or name
>
> Use the colon operator to index the first five observations of just the sepal  
> length and species from `iris`
>
> > ## Solution
> > Two options:
> > ```{r}
> > iris[1:5, c(1,5)]
> > iris[1:5,c("Sepal.Length","Species")]
> > ```
> {: .solution}
{: .challenge}

> ## Subsetting with Sequences
>
> Use the colon operator to index just the data on sepal size from `iris`
>
> > ## Solution
> > ```{r subset-sequence}
> > iris[, 1:2]
> > ```
> {: .solution}
{: .challenge}

> ## Adding a new variable
>
> We want to add a variable called "Petal.Color" the `iris` data frame to record a new set of 
> observations. Let's first define a new data frame 'iris.update' (so as not to modify our original raw data).
> 
> ```{r}
> iris.update = iris
> ```
> 
> Now, to initialize the variable, add a new character column to your data frame populated with
> no values to indicate that we have not recorded any observations.
> 
> > ## Solution
> > 
> > We have a couple of options:
> > 
> > 1) Define the vector and append it to the data frame using `cbind()`:
> > 
> > ```{r}
> > Petal.Color = character(length = dim(iris.update)[1]) # use the dim function to figure out how long to make the new vector
> > iris.update = cbind(iris.update, Petal.Color)
> > head(iris.update)
> > ```
> > 2) Directly populate the new column while creating it:
> > 
> > ```{r}
> > iris.update$Petal.Color = as.character("")
> > head(iris.update)
> > ```
> > 
> > There are also other ways to accomplish this task.
> {: .solution}
{: .challenge}

> ## Updating a Subset of Values
>
> Update the `iris.update` data frame by indicating that the "setosa" species had purple petals (without
> changing the values for the other species).
> 
> > ## Solution
> > ```{r}
> > iris.update[iris.update$Species == "setosa", ]$Petal.Color <- "purple"
> > iris.update
> > ```
> {: .solution}
{: .challenge}

***

{% include links.md %}
