---
title: "Lists and Advanced Indexing"
classdate: "9/2/2020 -- On Your Own"
teaching: 90
exercises: 30
questions:
- "What are lists and what is their relationship to vectors and data frames?"
- "How can we leverage indexing for more advanced data extraction?"
objectives:
- "Understand the structure and poperties of lists."
- "Be able to apply several techniques for extracting targeted data from data frames."
- "Combine different methods for accessing data with the assignment operator to update subsets of data."
keypoints:
- "Lists are a standard data structure in R in which each element can contain any other R object."
- "Lists can contain elements of different classes, unlike vectors."
- "Data frames are a specific type of list in which all elements are vectors of the same length. Each vector can contain data of different classes."
- "Use `object[[x]]` to select a single element from a list."
- "Each element of a list can be assigned a name that can be addressed using the `$` operator (e.g. `mylist$element1`)."
source: Rmd
---

```{r, include = FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-oyo-lists-")
```

##  On Your Own

### Lists

Lists in R act as containers. Unlike atomic vectors, the contents of a list are
not restricted to a single mode and can encompass any mixture of data
types. Lists are sometimes called "generic vectors"", because the elements of a
list can by of any type of R object, even lists containing further lists. This
property makes them fundamentally different from atomic vectors.

A list is a special type of vector. Each element can be a different type, and is not restricted to a single value.

Create lists using `list()` or coerce other objects using `as.list()`. An empty
list of the required length can be created using `vector()`

```{r}
x <- list(1, "a", TRUE, 1+4i)
x # A list does not print to the console like a vector. Instead, each element of the list starts on a new line.

x <- vector("list", length = 5) # empty list
length(x)
x 
```

&nbsp;

Vectors can be coerced to lists as follows:

```{r}
x <- 1:10
x <- as.list(x)
length(x)
```

&nbsp;
#### Indexing lists

Indexing works a bit differently for lists. The content of elements of a list can be retrieved by using double square brackets `[[n]]`, as opposed to the single square brackets `[n]` used for vectors and matrices.

```{r}
x[[1]]
```


> ## Examining Lists
>
> 1. What is the class of `x[1]`?
> 2. What is the class of `x[[1]]`?
>
> > ## Solution
> > 1. 
> >     ```{r examine-lists-1}
> >     class(x[1])
> >     ```
> > 2. 
> >     ```{r examine-lists-2}
> >     class(x[[1]])
> >     ```
> {: .solution}
{: .challenge}

&nbsp;

Elements of a list can be named (i.e. lists can have the `names` attribute)

```{r}
xlist <- list(a = "Karthik Ram", b = 1:10, data = head(iris))
xlist
names(xlist)
attributes(xlist)
```

&nbsp;

You can use the `$` operator to directly refer to named list elements by their name:

```{r}
xlist[[1]]
xlist$a
```

> ## Examining Named Lists
>
> 1. What is the length of the `xlist` object?
> 2. What is its structure?
>
> > ## Solution
> > 1. 
> >     ```{r examine-named-lists-1}
> >     length(xlist)
> >     ```
> > 2. 
> >     ```{r examine-named-lists-2}
> >     str(xlist)
> >     ```
> {: .solution}
{: .challenge} 

&nbsp;
#### Lists of lists!

A list can even have vectors (or other lists!) as one element of a list:

```{r}
x <- 1:10
y <- c(T, F, T, T) # "T" can be used in place of "TRUE"; "F" can be used in place of "FALSE"
z <- list(1, "a", TRUE, 1+4i)

my.list <- list(x, y, z)
my.list

my.list[[1]]
class(my.list[[1]])
class(my.list[[2]])

my.list <- list(x = x, y = y, z = x) # use the `=` to name your list elements
my.list$x
```

&nbsp;
#### Lists and functions

Lists can be extremely useful inside functions. Because the functions in R are 
able to return only a single object, you can "staple" together lots of different
kinds of results into a single object that a function can return.

Elements are indexed by double brackets. Double brackets return the object inside the indexed list element, while single brackets will just return a(nother) list containing the indicated elements. 

```{r}
my.list[1]
my.list[[1]]
```


&nbsp;

Again, if the elements of a list are named, they can be referenced by
the `$` notation (i.e. `xlist$data`). This can be useful if a function runs a complex analysis. It can export a lot of different types of information as different elements in a single list (raw data, processes data, analysis summary, analysis statistics, etc.)

***
### Data frames are specialized lists
At its heart, the data frame is a *special type of list* where every element of the list has same length (i.e. data frame is a "rectangular" list).

```{r}
dat <- data.frame(id = letters[1:10], x = 1:10, y = 11:20)
dat
```

&nbsp;

See that it is actually a special list:

```{r}
is.list(dat)
is.data.frame(dat) # "data.frame" is a sub-class of "list"
class(dat)
```


*** 
### Advanced indexing

Logical vectors can be created using `relational operators`:
* `<` = less than
* `>` = greater than
* `<=` = less than or equal to
* `>=` = greater than or equal to
* `==` = exactly equal to
* `!=` = not equal to
* `%in%` = is present in (to as if the value on the left is present in the vector/matrix on the right)

A few single variable examples:
```{r logical_variable_example}
1 == 1
1 == 2
1 != 1
4 > 7
18 %in% 1:10
18 %in% 15:25
```

&nbsp;

We can use these operators to query entire vectors and generate logical vectors:

```{r logical_vector_example}
# creating logical vectors from numeric data
x <- c(1, 2, 3, 11, 12, 13)
x < 10
x %in% 1:10
```

&nbsp;

We can use logical vectors to select data from a data frame.

```{r logical_vectors_indexing}
index <- iris$Species == 'setosa'
index
iris[index,]
```

&nbsp;

Often this operation is written as one line of code:

```{r logical_vectors_indexing2}
iris[iris$Species == 'setosa', ]
```

> ## Using logical indices
>
> Create a new data frame that is the subset of `iris` with sepal length greater than or
> equal to 5.0.
> 
> > ## Solution
> > ```{r}
> > iris.new = iris[iris$Sepal.Length > 5,]
> > ```
> {: .solution}
{: .challenge}

&nbsp;

In addition to the numeric comparisons, there are a set of operators that compare logical variables and output a new logical variable:
* `!` = NOT (changes `TRUE` to `FALSE` and vice versa)
* `&` = element-wise AND (both are true; outputs vector for vector input comparing elements)
* `&&` = logical AND (both are true; only considers first index of a vector)
* `|` = element-wise OR (one or both are true; outputs vector for vector input comparing elements)
* `||` = logical OR (both are true; only considers first index of a vector)

```{r}
truth = c(TRUE, FALSE, TRUE, TRUE)
lie = !truth
truth
lie

T & T
T & F
T | F
F | F
```


&nbsp;
#### Exercises

> ## Using logical indices
>
> Create a new data frame that is the subset of `iris` with sepal length greater than or > equal to 5.0 for the setosa species.
> 
> > ## Solution
> > ```{r}
> > iris.new = iris[iris$Sepal.Length > 5 & iris$Species == "setosa",]
> > ```
> {: .solution}
{: .challenge}

> ## Subsetting using a vector or name
>
> Use the colon operator to index the first five observations of just the sepal  
> length and species from `iris`
>
> > ## Solution
> > Two options:
> > ```{r}
> > iris[1:5, c(1,5)]
> > iris[1:5,c("Sepal.Length","Species")]
> > ```
> {: .solution}
{: .challenge}

> ## Subsetting with Sequences
>
> Use the colon operator to index just the data on sepal size from `iris`
>
> > ## Solution
> > ```{r subset-sequence}
> > iris[, 1:2]
> > ```
> {: .solution}
{: .challenge}

> ## Adding a new variable
>
> We want to add a variable called "Petal.Color" the `iris` data frame to record a new set of 
> observations. Let's first define a new data frame 'iris.update' (so as not to modify our original raw data).
> 
> ```{r}
> iris.update = iris
> ```
> 
> Now, to initialize the variable, add a new character column to your data frame populated with
> no values to indicate that we have not recorded any observations.
> 
> > ## Solution
> > 
> > We have a couple of options:
> > 
> > 1) Define the vector and append it to the data frame using `cbind()`:
> > 
> > ```{r}
> > Petal.Color = character(length = dim(iris.update)[1]) # use the dim function to figure out how long to make the new vector
> > iris.update = cbind(iris.update, Petal.Color)
> > head(iris.update)
> > ```
> > 2) Directly populate the new column while creating it:
> > 
> > ```{r}
> > iris.update$Petal.Color = as.character("")
> > head(iris.update)
> > ```
> > 
> > There are also other ways to accomplish this task.
> {: .solution}
{: .challenge}

> ## Updating a Subset of Values
>
> Update the `iris.update` data frame by indicating that the "setosa" species had purple petals (without
> changing the values for the other species).
> 
> > ## Solution
> > ```{r}
> > iris.update[iris.update$Species == "setosa", ]$Petal.Color <- "purple"
> > iris.update
> > ```
> {: .solution}
{: .challenge}

***

{% include links.md %}
